A Constructive Ordered Predicate Logic for Category Theory

### 2022-10-10 paper/paper.tex
intro/elim for cats and small cats
reformat figures

\begin{figure}
  \begin{center}
    \begin{tabular}{ |c|c| } 
      \hline
      Higher-Order Logic & Virtual Equipment Type Theory \\
      \hline
      Set $X$ & Category $\mathbb{C}$\\
      $X \times Y$ & $\mathbb C \times \mathbb D$ \\ 
      1 & $1$ \\
      $\mathcal P X$ & $\mathcal P^+ X, \mathcal P^- X$ \\
      $\{ (x,y) \in X \times Y | R(x,y) \}$ & $\sum_{\alpha:C;\beta:D} R$\\
      Function $f(x:X) : Y$ & Functor/Object $\alpha:\cat C \vdash A : \cat D$\\
      Relation $R(x,y)$ & Profunctor/Set $\alpha:\cat C; \beta:\cat D \vdash R$\\
      $R \wedge Q$ & $R \times Q$\\
      $\top$ & $1$\\
      $\forall x. P \Rightarrow Q$ & $\homrallXYtoZ {\alpha:\cat C} P Q \text{ and } \homlallXYtoZ {\alpha:\cat C} P Q$ \\
      $\exists x. P \wedge Q$ & $\tensorexistsXwithYandZ {\alpha:\cat C} P Q$\\
      $x =_X y$ & $\punitinXfromYtoZ {\cat C} \alpha \beta$\\
      Proof $\forall \overrightarrow \alpha. R_1 \wedge \cdots \Rightarrow Q$ & Nat. Trans./Element $\alpha_1,x_1: R_1(\alpha_1,\alpha_2),\ldots \vdash t : Q$\\
      \hline
    \end{tabular}
  \end{center}
  \caption{Analogy between Higher-Order Logic and \vett{} Judgments and Connectives}
\end{figure}

@@

After these forms we have types and terms, which represent the
meta-language that we use to talk about categories/profunctors/natural
transformations.  The types include the usual $\Pi$ and $\Sigma$ and
identity types of dependent type theory, along with universes of
categories, functors, profunctors and natural transformations.
%
Finally we have several forms of context which are used in the theory:
contexts $\Gamma$ of term variables with their types, contexts $\Xi$
of object variables with their categories, and contexts $\Phi$ of
element variables with their sets, alternating with those sets' object
variables.  A typical $\Phi$ will look like
\[
\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \alpha_2 : \cat C_2,
x_2:R_2(\alpha_2,\alpha_3), \alpha_3:\cat C_3, \ldots
\]
and represents the composition of the relations $R_1, R_2, R_3, \ldots,
R_n$.  We write $d^-(\Phi)$ for the first category variable in $\Phi$
(which we regard as the negative/contravariant position), $d^+(\Phi)$
for the last category variable in $\Phi$ (which we regard as the
positive/covariant position), and $\Phi_1 \jnctx \Phi_2$ for the append
of two category contexts where the last variable in $\Phi_1$ is equal to
the first variable in $\Phi_2$.  Formal inductive definitions are in the
appendix, but intuitively we have: 
\[
\begin{array}{ccl}
d^-(\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \ldots, x_n:R_n(\alpha_n,\alpha_n), \alpha_{n+1} : \cat C_{n+1}) & = & \alpha_1 : \cat C_1\\
d^+(\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \ldots, x_n:R_n(\alpha_n,\alpha_n), \alpha_{n+1} : \cat C_{n+1}) & = & \alpha_{n+1}:\cat C_{n+1}\\
(\Phi_1,\beta:\cat D) \jnctx (\beta:\cat D, \Phi_2) & = & \Phi_1, \beta : \cat D, \Phi_2
\end{array}
\]

@@

\begin{figure}
  \begin{mathpar}
    \begin{array}{rccl}
      \textrm{Categories} &\cat C, \cat D, \cat E & ::= & M \pipe \cat C \times \cat D \pipe \cat 1
      \pipe \graphProf \alpha \beta P \pipe \negPresheaf {\cat C} \pipe \posPresheaf {\cat C}
      \\
      \textrm{Objects} & a, b, c & ::= & \alpha \pipe M a \pipe (a,b) \pipe () \ldots\\
      \textrm{Sets} & P,Q,R & ::= & \punitinXfromYtoZ {\cat C} a b \pipe \tensorexistsXwithYandZ {\beta} P Q \pipe \homrallXYtoZ \beta P Q \pipe \homlallXYtoZ \alpha R S \\
      &&& 1 \pipe P \times Q\\
      &&& \pipe \harrapp M a b \pipe \negPresheafAppPtoX a b \pipe \posPresheafAppPtoX a b\\
      \textrm{Elements} & s,t,u & ::= & \ldots \\
      \\
      \textrm{Type} & A,B,C & ::= &
      \sum_{X:A} B \pipe \prod_{X:A} B \pipe \Id A M N \\
      &&& \Cat \pipe \varr{\cat C}{\cat D}\pipe \harr {\cat C} {\cat D}\pipe \pendallXdotY {\alpha:{\cat C}} R\\
      \textrm{Term} & L,M,N & ::= & X \pipe \lambda X:M. N \pipe M\,N \pipe (M,N) \pipe \pi_1 M \pipe \pi_2 M\\
      &&&\lceil \cat C \rceil \pipe \vlambda {\alpha:\cat C} a \pipe \hlambda {\alpha:\cat C}{\beta: \cat D} R \ldots \\
      \textrm{Type Context} & \Gamma, \Delta & ::= & \cdot \pipe \Gamma, X : A\\
      \textrm{Category Context} & \Xi, Z & ::= & \alpha:\cat C \pipe \alpha:\cat C;\beta: \cat D\\
      \textrm{Set Context} & \Phi,\Psi & ::= & \alpha : \cat C \pipe \Phi , x : P,\beta:\cat D\\
    \end{array}
  \end{mathpar}
  \caption{\vett{} Syntactic Forms}
  \label{fig:syntax}
\end{figure}

The \emph{formation conditions} or \emph{presuppositions} of a judgement
are the conditions under which a judgement ``type checks''.  We have
\begin{itemize}
\item ${\Gamma\vdash \cat C \isaCat}$ presupposes ${\Gamma \isadtctx}$
\item ${\Gamma \pipe \alpha : \cat C \vdash a : \cat D}$ presupposes
       $\Gamma \vdash \cat C\isaCat$ and $\Gamma \vdash \cat D \isaCat$.
\item ${\Gamma \pipe \Xi \vdash S \isaSet}$ presupposes
      $\Gamma \isadtctx$ and $\Gamma \vdash \Xi \boundary$
\item ${\Gamma \pipe \Phi \vdash s : R}$ presupposes
     ${\Gamma \vdash \Phi \isavectx}$
     and $\Gamma\pipe \underline \Phi \vdash R \isaSet$
\end{itemize}

%% Next, \cref{fig:formation} shows the \emph{formation} rules for these
%% various syntactic forms.

All other forms are typed with respect to a well-formed type context
$\Gamma$. Objects are typed additionally with an input object variable
$\alpha : \cat C$ and an output category $\cat D$; in the semantics
these objects are modeled as functors. A set $S$ is typed with respect
to a set context $\Xi$ to describe its covariant/contravariant
dependence on some input objects. These are semantically modeled as
profunctors. A transformation $s$ has a context of transformation
variable and a single set output. To be well formed, the context and set
must be parameterized by the same contravariant and covariant object
variables. To ensure this, we use a coercion operation $\underline \Phi$
that erases everything in the context but the left-most and right-most
object variables.

@@

In typical natural-deduction style, all syntactic forms typed in
context admit an action of substitution. For terms, this is the
ordinary type theoretic substitution. Objects $\alpha:\cat C
\vdash a : \cat D$ can be substituted for object variables $\beta :
\cat D$ in other objects. We can also substitute objects into
\emph{sets}, that is, if we have a set $P$ parameterized by a
contravariant variable $\alpha : \cat C$ and a covariant variable
$\beta: \cat D$, then we can substitute objects $a : \cat C$ and $b :
\cat D$ for these variables $P[a/\alpha;b/\beta]$. This generalizes the
ordinary precomposition of a relation by a function.  Semantically
this is the ``restriction'' of a profunctor along two functors, which
is just composition of functors if a profunctor is viewed as a functor
to $\Set$. Modeling this operation as a substitution considerably
simplifies reasoning using profunctors.
%
Finally we have the action of substitution on natural
transformations. A natural transformation is indexed by set variables
$x:R$, whose sets are each parameterized by a contravariant and a
covariant category variable $\alpha;\beta$. We can think of natural
transformations as \emph{polymorphic} in the categories involved, and
so when we make a transformation substitution, we substitute not just
the set variables, but we also \emph{instantiate} the polymorphic
category variables with objects.
%
The full syntactic details of substitution are included in the appendix.

@@

Recall that the basic entities of our type theory are categories,
objects/functors, sets/profunctors, and elements/natural
transformations.  In this section we discuss some connectives for
constructing categories, which are specified by introduction and
elimination rules (the $\beta\eta$ equality and substitution rules are
included in the appendix).  The introduction and elimination rules make
use of functors, profunctors, and natural transformations.
%
We introduce the connectives for defining categories in
Figure~\ref{fig:category-connectives}. 
%
First we introduce the additives: the unit category $1$ and product
category $\cat C \times \cat D$ have the usual introduction and
elimination rules defining functors to/from them.  
%
Next, we introduce the \emph{graph of a profunctor} $\graphProf \alpha
\beta P$. Just as a relation $R : A \times B \to \Set$ can be viewed
as a subset $\{ (a,b) \in A\times B | R(a,b)\}$, any profunctor $P :
C_-^o\times C_+ \to \Set$ can be viewed as a category with a functor
to $C_- \times C_+$ (no op), specifically a two-sided discrete
fibration. In set-based category theory, the objects are triples
$(a_-, a_+, s : P(a_-,a_+))$ and morphisms from $(a_-,a_+, s)$ to
$(a_-',a_+', s')$ are pairs of morphisms $f_- : a_- \to a_-'$ and $f_+
: a_+ \to a_+'$ such that $P(\id,f_+)(s) = P(f_-,\id)(s')$.
%
With various choices of $P$, this connective can be used to define the
arrow category, slice category and comma category constructions.
%
In our syntax we define it as the universal category $\cat C$ equipped
with functors to $\cat C_-$ and $\cat C_+$ and a morphism to $P$.

@@

\begin{figure}
  \begin{mathpar}
    \inferrule{ ~ }{\Gamma \vdash 1 \isaCat}
    \quad
    \inferrule*{ ~ }{\Gamma\pipe\alpha:C \vdash () : 1}\and

    \inferrule
        {\Gamma \vdash \cat C_1 \isaCat \and \cat C_2 \isaCat}
        {\Gamma \vdash {\cat C_1} \times {\cat C_2} \isaCat}
    \quad    
    \inferrule
    {\Gamma \pipe \alpha:\cat C \vdash a_1 : \cat C_1 \and \Gamma \pipe \alpha:\cat C \vdash a_2 : \cat C_2}
    {\Gamma \pipe\alpha:\cat C \vdash (a_1,a_2) : {\cat C_1} \times {\cat C_2}}
    \quad
    \inferrule
    {\Gamma \pipe\alpha:\cat C \vdash a : {\cat C_1} \times {\cat C_2}}
    {\Gamma\pipe\alpha:\cat C\vdash \pi_i a : \cat C_i}

    \inferrule
    {\Gamma \pipe \alpha:\cat C;\beta:\cat D \vdash P \isaSet}
    {\Gamma \vdash \graphProf{\alpha}{\beta} P \isaCat}
    \quad
    \inferrule
    {\defaultObCtx \vdash a_- : C_-\and
      \defaultObCtx \vdash a_+ : C_+\and
      \defaultObCtx \vdash s : P[a_-/\alpha;a_+/\beta]}
    {\defaultObCtx \vdash (a_-,a_+,s) : \graphProf{\alpha:\cat C_-}{\beta:\cat C_+} P}

    \inferrule
    {\defaultObCtx \vdash a : \graphProf{\alpha:\cat C_-}{\beta} P}
    {\defaultObCtx \vdash \pi_- a : \cat C_-}
    \quad
    \inferrule
    {\defaultObCtx \vdash a : \graphProf{\alpha}{\beta:C_+} P}
    {\defaultObCtx \vdash \pi_+ a : \cat C_+}
    \quad
    \inferrule
    {\defaultObCtx \vdash a : \graphProf{\alpha}{\beta} P}
    {\defaultObCtx \vdash \pi_e a : P[\pi_-a/\alpha;\pi_+a/\beta]}

    \inferrule
    {\Gamma \vdash \cat C \isaCat \and \cat C \isSmall}
    {\Gamma \vdash \negPresheaf {\cat C} \isaCat}
    \quad
    \inferrule
    {\Gamma \pipe d^-\Xi \vdash a : \cat C \and
     \Gamma \pipe d^+\Xi \vdash p : \negPresheaf {\cat C}
    }
    {\Gamma \pipe \Xi \vdash \negPresheafAppPtoX p a \isaSet}
    \quad
    \inferrule
    {\Gamma \pipe \alpha:\cat C;\beta:\cat D \vdash R : \isaSet}
    {\Gamma \pipe \beta:\cat D \vdash \lambda \alpha:\cat C. R : \negPresheaf {\cat C}}

    \inferrule
    {\Gamma \vdash \cat D \isaCat \and \cat D \isSmall}
    {\Gamma \vdash \posPresheaf {\cat D} \isaCat}
    \quad
    \inferrule
    {\Gamma \pipe d^-\Xi \vdash p : \posPresheaf {\cat D}\and
     \Gamma \pipe d^+\Xi \vdash a : \cat D
    }
    {\Gamma \pipe \Xi \vdash \posPresheafAppPtoX p a \isaSet}
    \quad
    \inferrule
    {\Gamma \pipe \alpha:\cat C;\beta:\cat D \vdash R : \isaSet}
    {\Gamma \pipe \alpha:\cat C \vdash \lambda \beta:\cat D. R : \posPresheaf {\cat D}}
  \end{mathpar}
  \caption{Category Conectives}
  \label{fig:category-connectives}
\end{figure}

@@

Next, in Figure~\ref{fig:set-connectives}, we cover the connectives for
the sets, i.e., constructions on profunctors.  In these constructions,
the sets/profunctors act like types, and they are given 
introduction and elimination rules for elements/natural
transformations.  
%
The top of the figure includes the formation rules for tensor products
$\tensorexistsXwithYandZ \beta P Q$, the covariant $\homrallXYtoZ
\alpha R P$ and contravariant homs $\homlallXYtoZ {\alpha} R P$ and
the unit type $\punitinXfromYtoZ {\cat C} {a_1}{a_2}$, and
the cartesian unit and product types $1$ and $R \times S$.  
%
The remainder of the figure gives the introduction and elimination rules
for these sets (the $\beta/eta$-rules are in the appendix).  
%
First, the unit set is our syntax for the profunctor of morphisms of a
category $\cat C$.
%
Its introduction and elimination rules are like those of an
\emph{ordered} variant of equality: the introduction rule is the
identity morphism (reflexivity) and the elimination rule is an
induction principle that is a variant of the J rule of equality
elimination: we can use a term of $s : \punitinXfromYtoZ {\cat C} a b$
by specifying the behavior when $s$ is of the form $\punitrefl \alpha$
in the form of a continuation $\alpha. t$. Next, the tensor product of
sets is a kind of combined existential quantifier and monoidal
product, which we combine into a single notation
$\tensorexistsXwithYandZ \beta P Q$, where $\beta$ is the covariant
variable of $P$ and the contravariant variable of $Q$. Then the
covariant variable of the tensor product is the covariant variable of
$Q$ and the contravariant variable similarly comes from $P$.
%
In ordinary category theory, this is the \emph{composition} of
profunctors, and is defined by a coend of a product. We require that
the variable $\beta$ quantifies over a small category $\cat D$, as in
general this composite doesn't exist for large categories.
%
The introduction and elimination are like those for a combined tensor
product and existential type: the introduction rule is a pair of
terms, with an appropriate instantiation of $\beta$, and the
elimination rule says to use a term of a tensor product, it is
sufficient to specify the behavior on two elements typed with an
arbitrary middle object $\beta$.

@@

\begin{figure}
  \begin{mathpar}
    % unit,

    \inferrule
    {\cat D \isSmall\\\\
      \Gamma \pipe d^-\Xi; \beta:D\vdash P \isaSet\\\\
     \Gamma \pipe \beta:D;d^+\Xi \vdash Q \isaSet}
    {\Gamma \pipe \Xi \vdash \tensorexistsXwithYandZ {\beta:D} P Q}

    \inferrule
    {d^-\Xi \isSmall\\\\
      \Gamma \pipe \alpha:\cat C; d^-\Xi\vdash R \\\\ \Gamma \pipe \alpha : \cat C; d^+\Xi\vdash P}
    {\Gamma \pipe \Xi \vdash \homlallXYtoZ {\alpha: \cat C} R P}

  \inferrule
  {d^+\Xi \isSmall\\\\ \Gamma \pipe d^+\Xi; \alpha : \cat C \vdash  R \\\\ \Gamma \pipe  d^-\Xi; \alpha : \cat C \vdash P}
  {\Gamma \pipe \Xi \vdash \homrallXYtoZ {\alpha:\cat C} R P}\\

  \inferrule
  {\Gamma \pipe d^-\Xi \vdash a_1 : \cat C\\\\ \Gamma \pipe d^+ \Xi \vdash a_2 : \cat C}
  {\Gamma \pipe \Xi \vdash \punitinXfromYtoZ {\cat C} {a_1} {a_2}}

    \inferrule{}{\Gamma\pipe \Xi \vdash 1 \isaSet}

>>>>

\begin{figure}
  \begin{center}
    \begin{tabular}{ |c|c| } 
      \hline
      Higher-Order Logic & Virtual Equipment Type Theory \\
      \hline
      Set $X$ & Category $\mathbb{C}$\\
      $X \times Y$ & $\mathbb C \times \mathbb D$ \\ 
      1 & $1$ \\
      $\mathcal P X$ & $\mathcal P^+ X$ \text{ and } $\mathcal P^- X$ \\
      $\{ (x,y) \in X \times Y | R(x,y) \}$ & $\sum_{\alpha:C;\beta:D} R$\\
      Function $f(x:X) : Y$ & Functor/Object $\alpha:\cat C \vdash A : \cat D$\\
      Relation $R(x,y)$ & Profunctor/Set $\alpha:\cat C; \beta:\cat D \vdash R$\\
      $R \wedge Q$ & $R \times Q$\\
      $\top$ & $1$\\
      $\forall x. P \Rightarrow Q$ & $\homrallXYtoZ {\alpha:\cat C} P Q \text{ and } \homlallXYtoZ {\alpha:\cat C} P Q$ \\
      $\exists x. P \wedge Q$ & $\tensorexistsXwithYandZ {\alpha:\cat C} P Q$\\
      $x =_X y$ & $\punitinXfromYtoZ {\cat C} \alpha \beta$\\
      Proof $\forall \overrightarrow \alpha. R_1 \wedge \cdots \Rightarrow Q$ & Nat. Trans./Element $\alpha_1,x_1: R_1(\alpha_1,\alpha_2),\ldots \vdash t : Q$\\
      \hline
    \end{tabular}
  \end{center}
  \caption{Analogy between Higher-Order Logic and \vett{} Judgments and Connectives}
\label{fig:hol-cats}
\end{figure}

@@

After these forms we have types and terms, which represent the
meta-language that we use to talk about categories/profunctors/natural
transformations.  The types include the usual $\Pi$ and $\Sigma$ and
identity types of dependent type theory, along with universes of
categories, functors, profunctors and natural transformations.

Finally we have several forms of context which are used in the theory.
The contexts $\Gamma$ of term variables with their types are as usual.
The category contexts $\Xi$ contain object variables with their
categories; the two forms of category context are $\alpha : \cat C$,
containing one variable that can be used both contravariantly and
covariantly, and $alpha : \cat C ; \beta : \cat D$, containing a
contravariant variable $\alpha$ and covariant variable $\beta$.
Finally, we have contexts $\Phi$ of
element variables with their sets, alternating with those sets' object
variables.  A typical $\Phi$ has the shape
\[
\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \alpha_2 : \cat C_2,
x_2:R_2(\alpha_2,\alpha_3), \ldots, R_n(\alpha_n,\alpha_{n+1}), \alpha_{n+1} : C_{n+1}
\]
and represents the composition of the relations $R_1, R_2, R_3, \ldots,
R_n$.  We write $d^-(\Phi)$ for the first category variable in $\Phi$
(which we regard as the negative/contravariant position), $d^+(\Phi)$
for the last category variable in $\Phi$ (which we regard as the
positive/covariant position), and $\Phi_1 \jnctx \Phi_2$ for the append
of two category contexts where the last variable in $\Phi_1$ is equal to
the first variable in $\Phi_2$.  Formal inductive definitions are in the
appendix, but intuitively we have: 
\[
\begin{array}{ccl}
d^-(\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \ldots, x_n:R_n(\alpha_n,\alpha_n), \alpha_{n+1} : \cat C_{n+1}) & = & \alpha_1 : \cat C_1\\
d^+(\alpha_1 : \cat C_1, x_1:R_1(\alpha_1,\alpha_2), \ldots, x_n:R_n(\alpha_n,\alpha_n), \alpha_{n+1} : \cat C_{n+1}) & = & \alpha_{n+1}:\cat C_{n+1}\\
(\Phi_1,\beta:\cat D) \jnctx (\beta:\cat D, \Phi_2) & = & \Phi_1, \beta : \cat D, \Phi_2
\end{array}
\]

@@

\begin{figure}[t]
  \begin{mathpar}
    \begin{array}{rccl}
      \textrm{Categories} &\cat C, \cat D, \cat E & ::= & M \pipe \cat C \times \cat D \pipe \cat 1
      \pipe \graphProf \alpha \beta P \pipe \negPresheaf {\cat C} \pipe \posPresheaf {\cat C}
      \\
      \textrm{Objects} & a, b, c & ::= & \alpha \pipe M a \pipe (a,b) \pipe () \ldots\\
      \textrm{Sets} & P,Q,R & ::= & \punitinXfromYtoZ {\cat C} a b \pipe \tensorexistsXwithYandZ {\beta} P Q \pipe \homrallXYtoZ \beta P Q \pipe \homlallXYtoZ \alpha R S \\
      &&& 1 \pipe P \times Q\\
      &&& \pipe \harrapp M a b \pipe \negPresheafAppPtoX a b \pipe \posPresheafAppPtoX a b\\
      \textrm{Elements} & s,t,u & ::= & \ldots \\
      \\
      \textrm{Type} & A,B,C & ::= &
      \sum_{X:A} B \pipe \prod_{X:A} B \pipe \Id A M N \\
      &&& \Cat \pipe \varr{\cat C}{\cat D}\pipe \harr {\cat C} {\cat D}\pipe \pendallXdotY {\alpha:{\cat C}} R\\
      \textrm{Term} & L,M,N & ::= & X \pipe \lambda X:M. N \pipe M\,N \pipe (M,N) \pipe \pi_1 M \pipe \pi_2 M \mid \text{refl}_{M} \\
      &&&\lceil \cat C \rceil \pipe \vlambda {\alpha:\cat C} a \pipe \hlambda {\alpha:\cat C}{\beta: \cat D} R \mid FIXME \ldots \\
      \textrm{Type Context} & \Gamma, \Delta & ::= & \cdot \pipe \Gamma, X : A\\
      \textrm{Category Context} & \Xi, Z & ::= & \alpha:\cat C \pipe \alpha:\cat C;\beta: \cat D\\
      \textrm{Set Context} & \Phi,\Psi & ::= & \alpha : \cat C \pipe \Phi , x : P,\beta:\cat D\\
    \end{array}
  \end{mathpar}
  \caption{\vett{} Syntactic Forms}
  \label{fig:syntax}
\end{figure}

Next, we overview our basic judgement forms along with the
\emph{formation conditions} or \emph{presuppositions} (the conditions
under which a judgement ``type checks'').  We have
\begin{itemize}
\item Categories: ${\Gamma\vdash \cat C \isaCat}$ presupposes ${\Gamma \isadtctx}$.

\item Objects/functors: ${\Gamma \pipe \alpha : \cat C \vdash a : \cat D}$ presupposes
  $\Gamma \vdash \cat C\isaCat$ and $\Gamma \vdash \cat D \isaCat$.
Objects are typed with an input object variable
$\alpha : \cat C$ and an output category $\cat D$; in the semantics
these objects are modeled as functors.

\item Sets/profunctors: ${\Gamma \pipe \Xi \vdash S \isaSet}$ presupposes
      $\Gamma \isadtctx$ and $\Gamma \vdash \Xi \boundary$
A set $S$ is typed with respect
to a set context $\Xi$ to describe its covariant/contravariant
dependence on some input objects. These are semantically modeled as
profunctors.

\item Elements/natural transformations: ${\Gamma \pipe \Phi \vdash s : R}$ presupposes
     ${\Gamma \vdash \Phi \isavectx}$
  and $\Gamma\pipe \underline \Phi \vdash R \isaSet$.
  A transformation $s$ has a context of transformation
variable and a single set output. To be well formed, the context and set
must be parameterized by the same contravariant and covariant object
variables. To ensure this, we use a coercion operation $\underline \Phi$
that erases everything in the context but the left-most and right-most
object variables, i.e. $\underline{\alpha : \cat C} = \alpha : \cat C$ and
$\underline{\Phi} = d^-(\Phi);d^+(\Phi)$ otherwise.

\item Meta-language types and terms: $\Gamma \vdash A \isaTy$ presupposes $\Gamma \isadtctx$ and $\Gamma
  \vdash M : A$ presupposes $\Gamma \isadtctx$ and $\Gamma \vdash A
  \isaTy$.
\end{itemize}

@@

All syntactic forms typed in context admit an action of
substitution. For (meta-language) types and terms, this is as
usual. Objects $\alpha:\cat C \vdash a : \cat D$ can be substituted for
object variables $\beta : \cat D$ in other objects. We can also
substitute objects into \emph{sets}, that is, if we have a set $P$
parameterized by a contravariant variable $\alpha : \cat C$ and a
covariant variable $\beta: \cat D$, then we can substitute objects $a :
\cat C$ and $b : \cat D$ for these variables $P[a/\alpha;b/\beta]$. This
generalizes the ordinary precomposition of a relation by a function.
Semantically this is the ``restriction'' of a profunctor along two
functors, which is just composition of functors if a profunctor is
viewed as a functor to $\Set$. Modeling this operation as a substitution
considerably simplifies reasoning using profunctors.
%
Finally we have the action of substitution on elements/natural transformations.
First, we can substitute elements/natural transformations for the set
variables in elements, denoting the composition of natural
transformations.  Second, an element is also parameterized
by a contravariant and a covariant category variable $\alpha;\beta$. We
can think of natural transformations as \emph{polymorphic} in the
categories involved, and so when we make a transformation substitution,
we also \emph{instantiate} the polymorphic category variables with
objects.
%
The full syntactic details of substitution are included in the appendix.

@@

In this section we discuss some connectives for constructing categories,
which are specified by introduction and elimination rules in
Figure~\ref{fig:category-connectives} (the $\beta\eta$ equality and
substitution rules are included in the appendix).  The introduction and
elimination rules make use of functors, profunctors, and natural
transformations.
%
First we introduce the additives: the unit category $1$ and product
category $\cat C \times \cat D$ have the usual introduction and
elimination rules defining functors to/from them.  
%
Next, we introduce the \emph{graph of a profunctor} $\graphProf \alpha
\beta P$. Just as a relation $R : A \times B \to \Set$ can be viewed
as a subset $\{ (a,b) \in A\times B | R(a,b)\}$, any profunctor $P :
C_-^o\times C_+ \to \Set$ can be viewed as a category with a functor
to $C_- \times C_+$ (no op), specifically a two-sided discrete
fibration. In set-based category theory, the objects are triples
$(a_-, a_+, s : P(a_-,a_+))$ and morphisms from $(a_-,a_+, s)$ to
$(a_-',a_+', s')$ are pairs of morphisms $f_- : a_- \to a_-'$ and $f_+
: a_+ \to a_+'$ such that $P(\id,f_+)(s) = P(f_-,\id)(s')$.
%
With various choices of $P$, this connective can be used to define the
arrow category, slice category and comma category constructions.
%
In our syntax we define it as the universal category $\cat C$ equipped
with functors to $\cat C_-$ and $\cat C_+$ and a morphism to $P$.

@@

\begin{figure}[t]
  \[
  \begin{array}{l}
    \text{Unit:} \quad
    \inferrule{ ~ }{\Gamma \vdash 1 \isaCat}
    \quad
    \inferrule*{ ~ }{\Gamma\pipe\alpha:C \vdash () : 1}\\\\

    \text{Product:} \quad
    \inferrule
        {\Gamma \vdash \cat C_1 \isaCat \and \cat C_2 \isaCat}
        {\Gamma \vdash {\cat C_1} \times {\cat C_2} \isaCat}
    \quad    
    \inferrule
    {\Gamma \pipe \alpha:\cat C \vdash a_1 : \cat C_1 \and \Gamma \pipe \alpha:\cat C \vdash a_2 : \cat C_2}
    {\Gamma \pipe\alpha:\cat C \vdash (a_1,a_2) : {\cat C_1} \times {\cat C_2}}
    \quad
    \inferrule
    {\Gamma \pipe\alpha:\cat C \vdash a : {\cat C_1} \times {\cat C_2}}
    {\Gamma\pipe\alpha:\cat C\vdash \pi_i a : \cat C_i}
    \\\\

    \text{Graph of a profuctor:} \\
    \inferrule
    {\Gamma \pipe \alpha:\cat C;\beta:\cat D \vdash P \isaSet}
    {\Gamma \vdash \graphProf{\alpha}{\beta} P \isaCat}
    \quad
    \inferrule
    {\defaultObCtx \vdash a_- : C_-\and
      \defaultObCtx \vdash a_+ : C_+\and
      \defaultObCtx \vdash s : P[a_-/\alpha;a_+/\beta]}
    {\defaultObCtx \vdash (a_-,a_+,s) : \graphProf{\alpha:\cat C_-}{\beta:\cat C_+} P}
    \\
    \inferrule
    {\defaultObCtx \vdash a : \graphProf{\alpha:\cat C_-}{\beta} P}
    {\defaultObCtx \vdash \pi_- a : \cat C_-}
    \quad
    \inferrule
    {\defaultObCtx \vdash a : \graphProf{\alpha}{\beta:C_+} P}
    {\defaultObCtx \vdash \pi_+ a : \cat C_+}
    \quad
    \inferrule
    {\defaultObCtx \vdash a : \graphProf{\alpha}{\beta} P}
    {\defaultObCtx \vdash \pi_e a : P[\pi_-a/\alpha;\pi_+a/\beta]}
    \\\\
    \text{Negative Presehaf:} \\
    \inferrule
    {\Gamma \vdash \cat C \isaCat \and \cat C \isSmall}
    {\Gamma \vdash \negPresheaf {\cat C} \isaCat}
    \quad
    \inferrule
    {\Gamma \pipe d^-\Xi \vdash a : \cat C \and
     \Gamma \pipe d^+\Xi \vdash p : \negPresheaf {\cat C}
    }
    {\Gamma \pipe \Xi \vdash \negPresheafAppPtoX p a \isaSet}
    \quad
    \inferrule
    {\Gamma \pipe \alpha:\cat C;\beta:\cat D \vdash R : \isaSet}
    {\Gamma \pipe \beta:\cat D \vdash \lambda \alpha:\cat C. R : \negPresheaf {\cat C}}
    \\
    \\
    \text{Positive Presehaf:} \\
    \inferrule
    {\Gamma \vdash \cat D \isaCat \and \cat D \isSmall}
    {\Gamma \vdash \posPresheaf {\cat D} \isaCat}
    \quad
    \inferrule
    {\Gamma \pipe d^-\Xi \vdash p : \posPresheaf {\cat D}\and
     \Gamma \pipe d^+\Xi \vdash a : \cat D
    }
    {\Gamma \pipe \Xi \vdash \posPresheafAppPtoX p a \isaSet}
    \quad
    \inferrule
    {\Gamma \pipe \alpha:\cat C;\beta:\cat D \vdash R : \isaSet}
    {\Gamma \pipe \alpha:\cat C \vdash \lambda \beta:\cat D. R : \posPresheaf {\cat D}}
  \end{array}
  \]
  \caption{Category Conectives}
  \label{fig:category-connectives}
\end{figure}

@@

Next, in Figure~\ref{fig:set-connectives}, we cover the connectives for
the sets/profunctors.  In these constructions, the sets/profunctors act
like types, and they are given introduction and elimination rules for
elements/natural transformations.
%
The top of the figure includes the formation rules for tensor products
$\tensorexistsXwithYandZ \beta P Q$, the covariant $\homrallXYtoZ
\alpha R P$ and contravariant homs $\homlallXYtoZ {\alpha} R P$ and
the unit type $\punitinXfromYtoZ {\cat C} {a_1}{a_2}$, and
the cartesian unit and product types $1$ and $R \times S$.  
%
The remainder of the figure gives the introduction and elimination rules
for these sets (the $\beta/eta$-rules are in the appendix).  
%
First, the unit set is our syntax for the profunctor of morphisms of a
category $\cat C$.
%
Its introduction and elimination rules are like those of an
\emph{ordered} variant of equality: the introduction rule is the
identity morphism (reflexivity) and the elimination rule is an
induction principle that is a variant of the J rule of equality
elimination: we can use a term of $s : \punitinXfromYtoZ {\cat C} a b$
by specifying the behavior when $s$ is of the form $\punitrefl \alpha$
in the form of a continuation $\alpha. t$. Next, the tensor product of
sets is a kind of combined existential quantifier and monoidal
product, which we combine into a single notation
$\tensorexistsXwithYandZ \beta P Q$, where $\beta$ is the covariant
variable of $P$ and the contravariant variable of $Q$. Then the
covariant variable of the tensor product is the covariant variable of
$Q$ and the contravariant variable similarly comes from $P$.
%
In ordinary category theory, this is the \emph{composition} of
profunctors, and is defined by a coend of a product. We require that
the variable $\beta$ quantifies over a small category $\cat D$, as in
general this composite doesn't exist for large categories.
%
The introduction and elimination are like those for a combined tensor
product and existential type: the introduction rule is a pair of
terms, with an appropriate instantiation of $\beta$, and the
elimination rule says to use a term of a tensor product, it is
sufficient to specify the behavior on two elements typed with an
arbitrary middle object $\beta$.

@@

\begin{figure}
    % unit,
  \[
  \begin{array}{l}
